---
title: "Computing PM2.5 health burden"
subtitle: "Example code with sample data"
author: "Meenakshi Kushwaha"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: flatly
    toc: true
    toc_float: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(here) # seting working directory
library(tidyverse) # tidy data analysis
```
### About
This file contains code for PM2.5 associated health impact analysis. The code uses sample exposure and outcome data that you can replace with your own dataset. Before starting your analysis please go throug the `README` file on background and instructions.

### Loading files

There are three files to be read into R and joined together. 

- Exposure data `exposure.csv` (`Sample Data` folder) 
- Health outcome data `outcome.csv` (`Sample Data` folder)
- Exposure response function `rr_gbd2019.csv` 

Below, we will read these files one by one, take a look at the data format and prepare them for joining together.

#### Exposure data

```{r read_exposure_data}
exposure_data <- read_csv(here("Sample Data", "exposure.csv" ))

knitr::kable(head(exposure_data), 
             align = 'c', 
             caption = "First six rows of exposure data")

exposure_join <- exposure_data %>% 
  mutate(pm25 = signif(pm25, 2)) # prepare file for joining by pm25 values
```

The sample data has annual mean PM2.5 `pm25` concentrations from five cities (city_A, city_B...) from 2016 to 2020. The dataset has `r nrow(exposure_data)` columns and `r ncol(exposure_data)` rows. 

To replace the `exposure.csv` file with your own exposure data. Make sure your data has the same file name and column names as the sample data. 

#### Health outcome data

```{r read_outcome_data}
outcome_data <- read_csv(here("Sample Data", "outcome.csv" ))

knitr::kable(head(outcome_data), 
             align = 'c', 
             caption = "First six rows of outcome data")

```

Annual mortality data from five cities (`city_A`, `city_B`...) from `2018` to `2020` for copd with following causes - diabetese, ihd, lung cancer and stroke. The dataset has `r nrow(outcome_data)` columns and `r ncol(outcome_data)` rows divided by age group and gender. 

To replace the `outcome.csv` file with your own health outcome data, ensure your data has the same file name and column names as the sample data. 


#### Exposure response function

This is PM2.5 exposure response relationship from GBD 2019 calculations. Source link ___?
More description - TBD
```{r read_exp_resp_table}
gbd_rr <- read_csv(here("rr_gbd2019.csv"))

knitr::kable(head(gbd_rr),
             align = "c",
             caption = "First six rows of GBD PM2.5 exposure response table")

gbd_rr_join <- gbd_rr %>% 
  mutate(exposure_spline = signif(exposure_spline, 2)) %>% 
  rename(pm25 = exposure_spline, 
         rr = mean,
         lcl = lower, 
         ucl = upper)

```


### Join files 

```{r file_joining}
exposure_rr <- left_join(exposure_join, # joining exposure data with exposure response table
                         gbd_rr_join, 
                         by = "pm25") # using pm25 column to join

exposure_rr <- exposure_rr %>% 
  mutate(paf_pm = (rr-1)/rr,
         lpaf_pm=(lcl-1)/lcl,
         upaf_pm=(ucl-1)/ucl)


outcome_exposure_rr <- left_join(outcome_data, # adding health outcome data
                                 exposure_rr, 
                                 by = c("year", "region", "age", "outcome")) 

```

### Calculate PM2.5 burden
PM2.5 health burdent is calculated by mulitplying the number of outcomes, for eg. number or `count` of deaths with the Population Attribuatable Fraction for PM2.5 `paf_pm`. Results can be aggregated by region and by year as below.

```{r}

pm25_burden <- outcome_exposure_rr %>% 
  mutate(pm25_burden = count*paf_pm)

# by region
pm25_burden_region <- pm25_burden %>% group_by(region, year) %>% summarize(tota_deaths = sum(pm25_burden, na.rm = TRUE))

# for year 2019
pm25_burden_2019 <- pm25_burden %>% filter(year == 2019) %>% group_by(region) %>%  summarize(total_deaths = sum(pm25_burden, na.rm = TRUE))

knitr::kable(pm25_burden_2019,
             caption = "Total deaths in 2019 attributable to PM2.5", 
             align = "c")

```

