---
title: "Computing PM2.5 health burden"
subtitle: "Example code with sample data"
author: "Meenakshi Kushwaha"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: flatly
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(here) # setting working directory
library(tidyverse) # tidy data analysis
```

### About

This file contains code for PM2.5 associated health impact analysis. The code uses sample exposure and outcome data that you can replace with your own dataset. Before starting your analysis please go throug the `README` file on background and instructions.

### Loading files

There are three main files to be read into R and joined together.

-   Exposure data `exposure.csv` (`Sample Data` folder)
-   Health outcome data `outcome.csv` (`Sample Data` folder)
-   Exposure response function `rr_gbd2019.csv`

Below, we will read these files one by one, take a look at the data format and prepare them for joining together.

#### Exposure data

```{r read_exposure_data}
exposure_data <- read_csv(here("Sample Data", "exposure.csv" ))

knitr::kable(head(exposure_data), 
             align = 'c', 
             caption = "First six rows of exposure data")

exposure_join <- exposure_data %>% 
  mutate(pm25 = signif(pm25, 2)) # prepare file for joining by pm25 values
```

The sample data has annual mean PM2.5 `pm25` concentrations from five cities (city_A, city_B...) from 2016 to 2020. The dataset has `r nrow(exposure_data)` columns and `r ncol(exposure_data)` rows.

To replace the `exposure.csv` file with your own exposure data. Make sure your data has the same file name and column names as the sample data.

#### Health outcome data

```{r read_outcome_data}
outcome_data <- read_csv(here("Sample Data", "outcome.csv" ))

knitr::kable(head(outcome_data), 
             align = 'c', 
             digits = 0,
             caption = "First six rows of outcome data")

```

Annual mortality data from five cities (`city_A`, `city_B`...) from `2018` to `2020` for copd with following causes - diabetese, ihd, lung cancer and stroke. The dataset has `r nrow(outcome_data)` columns and `r ncol(outcome_data)` rows divided by age group and gender.

To replace the `outcome.csv` file with your own health outcome data, ensure your data has the same file name and column names as the sample data.

#### Exposure response function

This is PM2.5 exposure response relationship from GBD 2019 calculations.^[Source:https://ghdx.healthdata.org/record/ihme-data/global-burden-disease-study-2019-gbd-2019-particulate-matter-risk-curves] 


```{r read_exp_resp_table}
gbd_rr <- read_csv(here("rr_gbd2019.csv"))

knitr::kable(head(gbd_rr),
             align = "c",
             digits = 3,
             caption = "First six rows of GBD PM2.5 exposure response table")

gbd_rr_join <- gbd_rr %>% 
  mutate(exposure_spline = signif(exposure_spline, 2)) %>% 
  rename(pm25 = exposure_spline, 
         rr = mean,
         lcl = lower, 
         ucl = upper)

```

### Join files

Combine exposure, outcome and exposure response data for finaly analysis. This includes following steps.

1.  Join exposure file with the exposure response table `exposure_rr`
2.  Add health outcome data `outcome_exposure_rr`.

For this step to work accurately, make sure that the joining columns or "key columns" are formatted similarly. Learn more in the `README` file.

```{r file_joining}
exposure_rr <- left_join(exposure_join, # joining exposure data with exposure response table
                         gbd_rr_join, 
                         by = "pm25") # using pm25 column to join

exposure_rr <- exposure_rr %>% 
  mutate(paf_pm = (rr-1)/rr,
         lpaf_pm=(lcl-1)/lcl,
         upaf_pm=(ucl-1)/ucl)


outcome_exposure_rr <- left_join(outcome_data, # adding health outcome data
                                 exposure_rr, 
                                 by = c("year", "region", "age", "outcome")) 

```

### Calculate total PM2.5 burden 

PM2.5 health burden is calculated by mulitplying the number of outcomes, for eg. `count` of deaths with the Population Attribuatable Fraction for PM2.5 `paf_pm`. Results can be aggregated by region or  by a particular year as below (2019, in this example). To generate results for a different year, replace the year value in `filter(year ==   )` call under the `# for year 2019` comment. 

```{r pm25_burden}

pm25_burden <- outcome_exposure_rr %>% 
  mutate(pm25_burden = count*paf_pm)

pm25_burden <- pm25_burden

# by region and year
pm25_burden_region <- pm25_burden %>% 
  group_by(region, year) %>% 
  summarize(deaths_total = sum(pm25_burden, na.rm = TRUE)) %>% 
  mutate(deaths_total = round(deaths_total, digits = 0))
  

write_csv(pm25_burden_region, here("Results","pm25_burden_region.csv"))


# for year 2019
pm25_burden_2019 <- pm25_burden %>% 
  filter(year == 2019) %>% 
  group_by(region) %>%  
  summarize(deaths_total = sum(pm25_burden, na.rm = TRUE)) %>% 
  mutate(deaths_total = round(deaths_total, digits = 0))

write_csv(pm25_burden_2019, here("Results","pm25_burden_2019.csv"))

knitr::kable(pm25_burden_2019,
             caption = "Total deaths in 2019 attributable to PM2.5", 
             align = "c")

```

### Calculate PM2.5 burden per 100K population

For this step, we need to combine the PM2.5 burden data with population of that year and then use the formula:   
Deaths per 100K = Total deaths/Population * 100000

```{r pm25_burden_per100k}
# Read and combine 2019 population data with the pm25_burden_2019 dataframe

pop_2019 <- read_csv(here("Sample Data", "population_2019.csv"))

pm25_burden_2019_per100k <- pm25_burden_2019 %>% 
  left_join(pop_2019, by = "region") %>% 
  mutate(deaths_per100k = deaths_total/pop_2019*100000) %>% 
  mutate(deaths_per100k = round(deaths_per100k, digits = 0))

write_csv(pm25_burden_2019_per100k, here("Results", "pm25_burden_2019_per100k.csv"))


knitr::kable(pm25_burden_2019_per100k,
             align = "c")

```

